
Notes from 10.09.21:

Read:
    Physics-informed  neural  networks:  A  deep  learning framework  for  solving  forward  and  inverse  problems  involving nonlinear  partial  differential  equations
    DEEPXDE: A DEEP LEARNING LIBRARY FOR SOLVING DIFFERENTIAL EQUATIONS
    Systems biology informed deep learning for inferring parameters and hidden dynamics

Code:
Verified that glycolysis.py and $fitzhugh_nagumo.py$ run.
There were some problems with version, SBINN uses an old version of deepxde, 0.13 instead of 0.10. Had to downgrade.
Led to some problems with dde.postprocessing.saveplot(). It didn't have the $output_dir$ paramter, but it said it did.

Started implementing a SBINN version for FitzHugh-Nagumo. 
Originally tried to just call the FitzHugh-Nagumo function instead of the glycolysis function. Obviously didn't work. 
Needed to adapt the FHN function to be compatible with the PINN function. Just copied the syntax form glycolysis.

Also needed to adapt the PINN function. Had to change a bunch of smaller stuff around:
    The variables at the start.
    The ODE function.
    Etc.
Had some problems with list shape - resolved.
The code doesn't use the GPU.
Was unsure about boundary conditions. I think it's a way/hack to get initial conditions. Was also unsure about how many is needed, settled on one for each variable in FHN, meaning 2. 
Was unsure about Observes. I think it's supposed to be the "raw" data. glycolysis had two, seemed to also work for FHN. 

Still allot of variables I'm unsure of. Probably need to test different ones. 
Still unsure about $feature_transform$ and $output_transform$.


To do:
Set up a conda environment. - done, or not. Doesn't look like it's working 
Fix gpu issue. Install gpu version of a module in the conda environment. - done
Maybe use a cluster as well. 
Create new version of dde.postprocessing.saveplot that supports output directory. - done
Figure out $observe_y4$ and 5, and also component = 4 and 5. Need to check the documentation. 
See if I can get the $model_restore_path$ thing working. - done




15.09.21:
Ran code for 5e5 epochs, got NaN after ~3e5 epochs.
Looks like the optimal value came quite early, after 17e3 epochs.

Did some small restructuring to get the $model_restore_path$ thing working.

